The Markov Link Method: a nonparametric approach to combine observations from multiple experiments

### 2018-08-07 paper/paper.tex
types typos typos

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $N^X$ and $N^Y$.  Here $N^X_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
\begin{enumerate}
    \item $\hat q(y|x)$ -- a point estimate for the link parameter $q^*(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $q^*(y|x)$ with high probability
\end{enumerate}
%
To produce these objects we proceeds as follows.  Let $\tilde P(\cdot |\ell) \sim \mathrm{Dirichlet}(N^X_\ell+1)$ and $\tilde H(\cdot |\ell) \sim \mathrm{Dirichlet}(N^Y_\ell+1)$ for each $\ell$.  For any pair $h,h'$ let 
\[
D(h|h') = \sum_{\ell,y} h(y|\ell) \log \frac{h(y|\ell)}{h'(y|\ell)}
\]
denote the Kullback-Leibler divergence from $h$ to $h'$.  For each $\tilde P,\tilde H$, let $E(\tilde P,\tilde H)$ denote the minimum value of $D(\tilde H|h')$ among all $h'$ such that there exists $q$ satisfying $h'(y|\ell)=\sum_x \tilde P(y|\ell)q(y|x)$.  Let $P,H$ denote variables carrying the distribution of $\tilde P,\tilde H$ after conditioned on $E(\tilde P,\tilde H)=0$.   Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$.  We then define the MLM objects as follows:
\begin{enumerate}
    \item $\hat q\triangleq\mathbb{E}[\argmin_{q\in \Theta(P,H)} \sum_{x,y} q(y|x)^2]$.
    \item The lower bound of $C_{x,y}$ is defined as the largest number $c$ such that $\min_{q\in \Theta(P,H)} q(y|x) \geq c$ with probability 97.5\%.  Likewise the upper bound is the smallest number $c$ such that $\max_{q\in \Theta(P,H)} q(y|x) \leq c$ with probability 97.5\%.
\end{enumerate} 
%
Code to compute these objects is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

Our approach arises naturally from a Bayesian point of view.  To understand them, first notice that there are several `nuisance parameters' we must consider in order to estimate the link $q^*$.  Let $p^*(x|\ell)$ denote the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$ and $h^*(y|\ell) \triangleq \sum_x p^*(x|\ell)q(y|x)$ denote the probability such a cell will have morphological type $y$.  Note that the distribution of the data $N^X,N^Y$ is entirely determined by $p^*,h^*$.   In particular, if someone changed $q^*$ without changing $p^*,h^*$ there would be no way for us to know.  If this can occur, $q^*$ is said to be `nonidentifiable.'   Therefore, instead of trying to directly estimate $q^*$ (which may be impossible to accurately estimate) we focus on three quantities which can be estimated from data:
%
\begin{align*}
q^*_{\mathrm{mid}} &\triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} q(y|x)^2\\
q^*_{\mathrm{lo},x,y} &\triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x) \\
q^*_{\mathrm{hi},x,y} &\triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x) 
\end{align*}
%
Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known; optimization details can be found in Appendix \ref{sec:optproblems}.  Moreover, regardless of any identifiability issues for $q^*$ itself, the objects $q^*_{\mathrm{lo},x,y}$ and $q^*_{\mathrm{hi},x,y}$ are always identifiable for every $(x,y)$ and define bounds on the true value of $q^*(y|x)$.

>>>>

In the specific case we will focus on, all of the experimental data can be summarized with two matrices, $N^X$ and $N^Y$.  Here $N^X_{\ell x}$ indicates the number of specimens gathered with procedure $\ell$ that have transcriptomic type $x$.  Likewise $N^Y_{\ell y}$ indicates the number of specimens gathered with procedure $\ell$ that have morphological type $y$.  For each transcriptomic type $x$ and morphological type $y$, the Markov Link Method uses the data to produce two objects:
%
\begin{enumerate}
    \item $\hat q(y|x)$ -- a point estimate for the link parameter $q^*(y|x)$
    \item $C_{x,y}$ -- a credible interval which contains the true link parameter $q^*(y|x)$ with high probability
\end{enumerate}
%
To produce these objects we proceeds as follows.  Let $\tilde P(\cdot |\ell) \sim \mathrm{Dirichlet}(N^X_\ell+1)$ and $\tilde H(\cdot |\ell) \sim \mathrm{Dirichlet}(N^Y_\ell+1)$ for each $\ell$.  For any pair $h,h'$ let 
\[
D(h|h') = \sum_{\ell,y} h(y|\ell) \log \frac{h(y|\ell)}{h'(y|\ell)}
\]
denote the Kullback-Leibler divergence from $h$ to $h'$.  For each $\tilde P,\tilde H$, let $E(\tilde P,\tilde H)$ denote the minimum value of $D(\tilde H|h')$ among all $h'$ such that there exists $q$ satisfying $h'(y|\ell)=\sum_x \tilde P(y|\ell)q(y|x)$.  Let $P,H$ denote variables carrying the distribution of $\tilde P,\tilde H$ after conditioning on $E(\tilde P,\tilde H)=0$.   Let $\Theta(p,h)=\{q:\ h(y|\ell) = \sum_x p(x|\ell)q(y|x)\}$.  We then define the MLM objects as follows:
\begin{enumerate}
    \item $\hat q\triangleq\mathbb{E}[\argmin_{q\in \Theta(P,H)} \sum_{x,y} q(y|x)^2]$.
    \item The lower bound of $C_{x,y}$ is defined as the largest number $c$ such that $\min_{q\in \Theta(P,H)} q(y|x) \geq c$ with probability 97.5\%.  Likewise the upper bound is the smallest number $c$ such that $\max_{q\in \Theta(P,H)} q(y|x) \leq c$ with probability 97.5\%.
\end{enumerate} 
%
Code to compute these objects is published at https://github.com/jacksonloper/markov-link-method, including a tutorial-style ipython notebook detailing every computation made in this paper.

Our approach arises naturally from a Bayesian point of view.  To understand how, first notice that there are several `nuisance parameters' we must consider in order to estimate the link $q^*$.  Let $p^*(x|\ell)$ denote the probability that a cell sampled with procedure $\ell$ will have transcriptomic type $x$ and $h^*(y|\ell) \triangleq \sum_x p^*(x|\ell)q(y|x)$ denote the probability such a cell will have morphological type $y$.  Note that the distribution of the data $N^X,N^Y$ is entirely determined by $p^*,h^*$.   In particular, if someone changed $q^*$ without changing $p^*,h^*$ there would be no way for us to know.  If this can occur, $q^*$ is said to be `nonidentifiable.'   Therefore, instead of trying to directly estimate $q^*$ (which may be impossible to accurately estimate) we focus on three quantities which can be estimated from data:
%
\begin{align*}
q^*_{\mathrm{mid}} &\triangleq \argmin_{q\in \Theta(p^*,h^*)} \sum_{x,y} q(y|x)^2\\
q^*_{\mathrm{lo},x,y} &\triangleq \min_{q\in \Theta(p^*,h^*)} q(y|x) \\
q^*_{\mathrm{hi},x,y} &\triangleq \max_{q\in \Theta(p^*,h^*)} q(y|x) 
\end{align*}
%
Note that all three of these optimization problems have unique solutions and would be straightforward to solve if $p^*,h^*$ were known; optimization details can be found in Appendix \ref{sec:optproblems}.  Moreover, regardless of any identifiability issues for $q^*$ itself, the objects $q^*_{\mathrm{lo},x,y}$ and $q^*_{\mathrm{hi},x,y}$ are always identifiable for every $(x,y)$ and define bounds on the true value of $q^*(y|x)$.
